---
title: "SPAM/HAM classification using caret (from Ch. 4 of 'Machine Learning with R')"
author: "Jesus M. Castagnetto"
date: '2015-01-01'
output:
    html_document:
        theme: readable
        keep_md: true
        toc: true
---

```{r echo=FALSE}
library(knitr)
opts_chunk$set(cache=TRUE, comment="", warning=FALSE, message=FALSE)
```

## Background

### Motivation

I am currently reading the book "Machine Learning with R"[^mlr] by Brent Lantz,
and also want to learn more about the `caret`[^caret] package, so I decided to replicate
the SPAM/HAM classification example from the chapter 3 of the book using `caret` instead of the 
`e1071`[^e1071] package used in the text.

[^mlr]: [Book page at Packt](https://www.packtpub.com/big-data-and-business-intelligence/machine-learning-r)

[^caret]: [The caret package](http://caret.r-forge.r-project.org/) site

[^e1071]: [http://cran.r-project.org/web/packages/e1071/index.html](http://cran.r-project.org/web/packages/e1071/index.html)

### Preliminary information

The dataset used in the book is a modified version of the "SMS Spam Collection v.1" created by Tiago A. Almeida and José Maria Gómez Hidalgo[^smsspamcoll],
as described in Chapter 4 ("*Probabilistic Learning -- Clasification Using Naive Bayes") of the aforementioned book. 

[^smsspamcoll]: [http://www.dt.fee.unicamp.br/~tiago/smsspamcollection/](http://www.dt.fee.unicamp.br/~tiago/smsspamcollection/)

You can get the modified dataset from the book's page at Packt, but be aware
that you will need to register to get the files. If you rather don't do that,
you can get the original data files from the original creator's site.

To simplify things, we are going to use the original dataset.

For this excercise we will use the `caret` package to do the kNN modeling and 
prediction, the `tm` package to generate the text corpus,
the `pander` package to be able to output nicely formated tables, 
and the `doMC` to take advantage of parallel processing with multiple cores.
Also, we will define some utility functions to simplify matters later in the code.

```{r warning=FALSE, message=FALSE}
# libraries needed by caret
library(klaR)
library(MASS)
# for the Naive Bayes modelling
library(caret)
# to process the text into a corpus
library(tm)
# to get nice looking tables
library(pander)
panderOptions("table.style", "rmarkdown")
# to simplify selections
library(dplyr)
library(doMC)
registerDoMC(cores=4)

# a utility function for % freq tables
frqtab <- function(x, caption) {
    round(100*prop.table(table(x)), 1)
}
# utility function to summarize model comparison results
sumpred <- function(cm) {
    summ <- list(TN=cm$table[1,1],  # true negatives
                 TP=cm$table[2,2],  # true positives
                 FN=cm$table[1,2],  # false negatives
                 FP=cm$table[2,1],  # false positives
                 acc=cm$overall["Accuracy"],  # accuracy
                 sens=cm$byClass["Sensitivity"],  # sensitivity
                 spec=cm$byClass["Specificity"],  # specificity
                 PPV=cm$byClass["Pos Pred Value"], # positive predictive value
                 NPV=cm$byClass["Neg Pred Value"]) # negative predictive value
    lapply(summ, FUN=round, 2)
}
```

## Reading and preparing the data

We start by downloading the zip file with the dataset, and reading the file into
a dataframe. We then assign the appropiate names to the columns, and convert
the `type` into a factor. Finally, we randomize the data frame.

```{r}
if (!file.exists("smsspamcollection.zip")) {
download.file(url="http://www.dt.fee.unicamp.br/~tiago/smsspamcollection/smsspamcollection.zip",
              destfile="smsspamcollection.zip", method="curl")
}
sms_raw <- read.table(unz("smsspamcollection.zip","SMSSpamCollection"),
                      header=FALSE, sep="\t", quote="", stringsAsFactors=FALSE)
colnames(sms_raw) <- c("type", "text")
sms_raw$type <- factor(sms_raw$type)
# randomize it a bit
set.seed(12358)
sms_raw <- sms_raw[sample(nrow(sms_raw)),]
str(sms_raw)
```

The modified data used in the book has 5559 SMS messages, whereas the original
data used here has `r nrow(sms_raw)` rows.

According to the documentation that accompanies the data file, 86.6% of the
entries correspond to legitimate messages ("ham"), and 13.4% to spam messages.

```{r results='asis'}
ft_orig <- frqtab(sms_raw$type)
pander(ft_orig, caption="Original message type frequencies (%)")
```

## Preparing the data

We wil proceed in a similar fashion as described in the book, but make use
of `dplyr` syntax to concatenate the text cleanup/transformation operations

```{r}
sms_corpus <- Corpus(VectorSource(sms_raw$text))
sms_corpus_clean <- sms_corpus %>%
    tm_map(content_transformer(tolower)) %>% 
    tm_map(removeNumbers) %>%
    tm_map(removeWords, stopwords()) %>%
    tm_map(removePunctuation) %>%
    tm_map(stripWhitespace)
sms_dtm <- DocumentTermMatrix(sms_corpus_clean)
```

## Creating a classification model witn Naive Bayes

### Generating the training and testing data sets

```{r results='asis'}
train_index <- createDataPartition(sms_raw$type, p=0.75, list=FALSE)
sms_raw_train <- sms_raw[train_index,]
sms_raw_test <- sms_raw[-train_index,]
sms_corpus_clean_train <- sms_corpus_clean[train_index]
sms_corpus_clean_test <- sms_corpus_clean[-train_index]
sms_dtm_train <- sms_dtm[train_index,]
sms_dtm_test <- sms_dtm[-train_index,]
ft_df <- as.data.frame(cbind(ft_orig, frqtab(sms_raw_train$type), frqtab(sms_raw_test$type)))
colnames(ft_df) <- c("Original", "Training set", "Test set")
pander(ft_df,
       caption=paste0("Comparison of SMS type frequencies among datasets"))
```

Select terms that appear at least 5 times in the training document term matrix

```{r}
sms_dict <- findFreqTerms(sms_dtm_train, lowfreq=5)
sms_train <- DocumentTermMatrix(sms_corpus_clean_train, list(dictionary=sms_dict))
sms_test <- DocumentTermMatrix(sms_corpus_clean_test, list(dictionary=sms_dict))
```

Convert values in the term matrix into factors

```{r}
# modified sligtly fron the code in the book
convert_counts <- function(x) {
    x <- ifelse(x > 0, 1, 0)
    x <- factor(x, levels = c(0, 1), labels = c("Absent", "Present"))
}
sms_train <- sms_train %>% apply(MARGIN=2, FUN=convert_counts)
sms_test <- sms_test %>% apply(MARGIN=2, FUN=convert_counts)
```

Training the model

```{r warning=FALSE, message=FALSE}
ctrl <- trainControl(method="cv", 10)
set.seed(12358)
sms_model1 <- train(sms_train, sms_raw_train$type, method="nb",
                trControl=ctrl)
sms_model1
sms_predict1 <- predict(sms_model1, sms_test)
cm1 <- confusionMatrix(sms_predict1, sms_raw_test$type, positive="spam")
cm1

set.seed(12358)
sms_model2 <- train(sms_train, sms_raw_train$type, method="nb", 
                    tuneGrid=data.frame(.fL=1, .usekernel=FALSE),
                trControl=ctrl)
sms_model2
sms_predict2 <- predict(sms_model2, sms_test)
cm2 <- confusionMatrix(sms_predict2, sms_raw_test$type, positive="spam")
cm2
```

```{r echo=FALSE}
#save(sms_raw, sms_raw_test, sms_raw_train,
#     sms_corpus, sms_corpus_clean, sms_corpus_clean_test,
#     sms_corpus_clean_train, sms_dict, sms_dtm, sms_dtm_test,
#     sms_dtm_train, sms_model1, sms_model2, sms_predict1,
#     sms_predict2, cm1, cm2, file="smsobjs.Rdata")
```


```{r results='asis'}
# from the table on page 115 of the book
tn=1203
tp=151
fn=32
fp=4
book_example1 <- list(
    TN=tn,
    TP=tp,
    FN=fn,
    FP=fp,
    acc=(tp + tn)/(tp + tn + fp + fn),
    sens=tp/(tp + fn),
    spec=tn/(tn + fp),
    PPV=tp/(tp + fp),
    NPV=tn/(tn + fn))

# from the table on page 116 of the book
tn=1204
tp=152
fn=31
fp=3
book_example2 <- list(
    TN=tn,
    TP=tp,
    FN=fn,
    FP=fp,
    acc=(tp + tn)/(tp + tn + fp + fn),
    sens=tp/(tp + fn),
    spec=tn/(tn + fp),
    PPV=tp/(tp + fp),
    NPV=tn/(tn + fn))

b1 <- lapply(book_example1, FUN=round, 2)
b2 <- lapply(book_example2, FUN=round, 2)
m1 <- sumpred(cm1)
m2 <- sumpred(cm2)
model_comp <- as.data.frame(rbind(b1, b2, m1, m2))
rownames(model_comp) <- c("Book model 1", "Book model 2", "Caret model 1", "Caret model 2")
pander(model_comp, split.tables=Inf, keep.trailing.zeros=TRUE,
       caption="Model results when comparing predictions and test set")
```


sensitivy is also known as *recall*, PPV is also know as *precision*

## Applying the model to a different SMS SPAM data set

We will use "The SMS Spam Corpus v.0.1 Big"[^smsspamcorpus]
by José María Gómez Hidalgo and Enrique Puertas Sanz, which contains a
total of 1002 legitimate messages and a total of 322 spam messages.

[^smsspamcorpus]: [http://www.esp.uem.es/jmgomez/smsspamcorpus](http://www.esp.uem.es/jmgomez/smsspamcorpus)

The data file has inconsistent formatting, so we need to clean it up a bit
before using it.

```{r}
library(plyr)
smsdata <- readLines(unz("SMSSpamCorpus01.zip", "english_big.txt")) %>%
    iconv(from="latin1", to="UTF-8") 
smsdata <- gsub(",(spam|ham)$", "|\\1", smsdata)
smsdata <- strsplit(smsdata, "|", fixed=TRUE)
smsdata <- ldply(smsdata)
colnames(smsdata) <- c("text", "type")
smsdata$type <- factor(smsdata$type)
sms_corpus2 <- Corpus(VectorSource(smsdata$text))
sms_corpus2_clean <- sms_corpus2 %>%
    tm_map(content_transformer(tolower)) %>% 
    tm_map(removeNumbers) %>%
    tm_map(removeWords, stopwords()) %>%
    tm_map(removePunctuation) %>%
    tm_map(stripWhitespace)

sms_dtm2 <- DocumentTermMatrix(sms_corpus2_clean, list(dictionary=sms_dict))
sms_test2 <- sms_dtm2 %>% apply(MARGIN=2, FUN=convert_counts)
sms_predict3 <- predict(sms_model1, sms_test2)
cm3 <- confusionMatrix(sms_predict3, smsdata$type, positive="spam")
cm3

sms_predict4 <- predict(sms_model2, sms_test2)
cm4 <- confusionMatrix(sms_predict4, smsdata$type, positive="spam")
cm4
```

Surprisingly well


```{r results='asis'}
m3 <- sumpred(cm3)
m4 <- sumpred(cm4)
model_comp <- as.data.frame(rbind(m3,m4))
rownames(model_comp) <- c("Caret model 1", "Caret model 2")
pander(model_comp, split.tables=Inf, keep.trailing.zeros=TRUE,
       caption="Applying the caret models to a new data set")
```

## Reproducibility information

The dataset used is original "SMS Spam Collection v.1" by Tiago A. Almeida and José Maria Gómez Hidalgo[^smsspamcoll], 
as described in the chapter 4 of book "Machine Learning with R" by Brett Lantz 
(ISBN 978-1-78216-214-8).

```{r}
sessionInfo()
```
